Some examples in extended logic counting classes

### 2016-12-28 LICS/counting.tex
27 dic
Primera versión de Sección 4.

Assume that $\alpha$ is a sentence in $\qso$. Then slightly abusing notation, we also use $\alpha$ to denote a function defined over $\R$ such that for every $\A \in \ostr[\R] $:
\begin{eqnarray*}
	\alpha(\enc(\A)) & = & \sem{\alpha}(\A).
\end{eqnarray*}
\begin{definition} \label{cap-def}
	Let $\FF$ a fragment of $\qso$ and $\CC$ a function complexity class. Then $\FF$ {\em captures $\CC$ over ordered structures} if

	\begin{itemize}
	\item[(1)] for every $\alpha$ in $\FF$ over a signature $\R$. There exists $f \in \CC$ such that for every $\A \in \ostr[\R]$:  $\alpha(\enc(\A)) = f(\enc(\A))$; and

	\item[(2)] for every $f \in \CC$, let $f$ be defined over $\R$. There exists $\alpha$ in $\FF$ such that for every $\A \in \ostr[\R]$:  $f(\enc(\A)) = \alpha(\enc(\A))$.
	\end{itemize} 
\end{definition}
%In the previous definition, if $\KK$ is the class of all finite structures, then we say that $\FF$ captures $\CC$ over the class of all structures, and if $\KK$is the class of all ordered finite structures, then we say that $\FF$ captures $\CC$ over the class of ordered structures.

\begin{theorem} \label{captfp}
	$\qfo(\lfp)$ captures $\fp$ over ordered structures.
\end{theorem}

\begin{theorem} \label{qfo-pfp-cap}
	$\qfo(\pfp)$ captures $\nfpspace$ over ordered structures.
\end{theorem}

Saluja et. al \cite{SalujaST95} define a family of counting classes $\#\LL$ for each fragment $\LL$ of $\so$. For a formula $\varphi(x,X)$ over $\R$, the function $f_{\varphi(x,X)}$ is defined as
\[
f_{\varphi(x,X)}(\A) = \vert \{\langle e,P\rangle\mid \A\models\varphi(e,P)\}\vert.
\]
for each $\A\in\ostr[\R]$. A function $f:\Sigma^*\to\nat$ is in $\#\LL$ if there exists an $\LL$ formula $\varphi(x,X)$ such that $f = f_{\varphi(x,X)}$. Let $\sqso(\LL)$ be a fragment of $\qso$ defined by the following grammar:
\[
\alpha := \sa{x}\varphi \ \mid \ \sa{X} \varphi \ \mid \ \sa{x}\alpha \ \mid \ \sa{X}\alpha,
\]
where $\varphi$ is an $\LL$-formula. It can be seen that $\sqso(\LL)$ captures $\#\LL$ over ordered structures and that $\sqso(\LL) \subseteq \eqso(\LL)$.

\begin{theorem} \label{eqso-fo-cap}
	$\eqso(\fo)$ captures $\shp$ over ordered structures.
\end{theorem}

\begin{corollary} \label{eqso-lfp-cap}
	$\eqso(\lfp)$ captures $\shp$ over ordered structures.
\end{corollary}

\begin{theorem} \label{eqso-eso-cap}
	$\eqso(\eso)$ captures $\spp$ over ordered structures.
\end{theorem}

\begin{theorem} \label{qso-pfp-cap}
	$\qso(\pfp)$ captures $\fpspace$ over ordered structures.
\end{theorem}

>>>>

In this section, we show that by syntactically restricting $\qso$ one can capture different function complexity classes. 
In other words, by using $\qso$ we can extend the theory of descriptive complexity~\cite{immerman2012descriptive} from decision problems to counting problems. 
For this, we first formalize the notion of \emph{capturing} a function complexity class to then show how to capture classes like $\shp$, $\fp$, and $\fpspace$.

Fix a relational signature $\R = \{R_1, \ldots, R_k\}$ and assume that $\A$ is an ordered finite $\R$-structure with a domain $A$ containing $n$ elements. Recall that  $<$ is a linear order on $A$, say $a_1 < a_2 < \ldots < a_n$. For every $i \in \{1, \ldots, k\}$, define the encoding of $R_i^\A$, denoted by $\enc(R_i^\A)$, as the following binary string. Assume that $\ell = \arity(R_i)$ and consider an enumeration of the $\ell$-tuples over $A$ in the lexicographic order induced by $<$. 
%(that is, $(a_1, \ldots, a_1, a_1)$, $(a_1, \ldots, a_1, a_2)$, $\ldots$, $(a_n, \ldots, a_n, a_{n-1})$, $(a_n, \ldots, a_n, a_n)$). 
Then let $\enc(R_i^\A)$ be a binary string of length $n^\ell$ such that the $i$-th bit of $\enc(R_i^\A)$ is 1 if the $i$-th tuple in the previous enumeration belongs to $R_i^\A$, and 0 otherwise. Moreover, define the encoding of $\A$, denoted by $\enc(\A)$, as the following binary string~\cite{L04}:
\begin{eqnarray*}
	\enc(\A) & = & 0^n \, 1 \, \enc(R_1^\A) \, \cdots \, \enc(R_k^\A).
\end{eqnarray*}
%We define the class of all $\R$-functions, denoted by $\Func(\R)$, as the class of all functions $f: \ostr \rightarrow \bbN$.
%Given a function complexity class $\CC$ (i.e. $f: \Sigma^* \rightarrow \bbN$ for every $f \in \CC$), we say that a function $f \in \Func(\R)$ can be computed in $\CC$ if there exists $g \in \CC$ such that $f(\A) = g(\enc(\A))$ for every $\A \in \ostr$. 
%Note that the function $g$ outputs $f$ for encodings of structures and can behave arbitrarily otherwise.
We can now formalize the notion of capturing a function complexity class over $\R$-structures.
\begin{definition} \label{cap-def}
	Let $\FF$ be a fragment of $\qso$ and $\CC$ a function complexity class. Then we say that {\em  $\FF$ captures $\CC$ over $\R$-structures} if the  following conditions are true:
	\begin{enumerate}
		\item for every $\alpha \in \FF$, there exists $f \in \CC$ such that $\sem{\alpha}(\A) = f(\enc(\A))$ for every $\A \in \ostr[\R]$.

		\item for every $f \in \CC$, there exists $\alpha \in \FF$ such that   $f(\enc(\A)) = \sem{\alpha}(\A)$ for every $\A \in \ostr[\R]$.
	\end{enumerate} 
	Finally, we say that $\FF$ captures $\CC$ if $\FF$ captures $\CC$ over $\R$-structures for every relational signature $\R$.
\end{definition}
%For the sake of simplification, we denote the first condition by $\FF \subseteq \CC$ and the second condition by $\CC \subseteq \FF$.
Note that in Definition~\ref{cap-def} the function $f \in \CC$ and the formula $\alpha \in \FF$ must coincide in all the strings that encode $\R$-structures. Since we want to capture $\CC$ over a fix set of structures (e.g. graphs, matrices), it is natural to just consider strings that encodes $\R$-structures. 
This definition is fairly standard in descriptive complexity \cite{L04,immerman2012descriptive} and all notions for capturing complexity classes restrict $f \in \CC$ similarly.

Then, what functional complexity classes can be captured with fragments of $\qso$?
For answering this question, it is reasonable to start with $\shp$ who is one of the most well-known functional complexity class~\cite{arora2009computational}. 
Since $\shp$ has a strong similarity with $\np$, one could expect a ``Fagin-like'' Theorem~\cite{fagin1974generalized} for this class. 
Actually, in~\cite{SalujaST95} it was shown that the class $\sfo$ captures $\shp$.
In our setting, the class $\sfo$ is contained in $\eqso(\fo)$ which also captures $\shp$ as expected. 
\begin{proposition} \label{eqso-fo-cap}
	$\eqso(\fo)$ captures $\shp$ over ordered structures.
\end{proposition}
Recall that (see Section~\ref{sec:previous}) that every function class $\# \LL$ is contained in $\eqso(\LL)$. Thus, by~\cite{SalujaST95} it directly follows that every $\shp$-function can be defined in $\eqso(\LL)$. The other direction is also direct and follows by showing that $\shp$ is closed under first- and second-order sum.
Moreover, these closure properties also hold if we change the boolean core $\fo$ to any polynomial-time logic like $\lfp$. This gives the following corollary.
\begin{corollary} \label{eqso-lfp-cap}
	$\eqso(\lfp)$ captures $\shp$ over ordered structures.
\end{corollary}
By following the same approach as~\cite{SalujaST95}, Compton et al.~\cite{ComptonG96} show that $\seso$ captures $\spp$. As one could expect, if we parametrize $\eqso$ with $\eso$, we can also capture~$\spp$.
\begin{proposition} \label{eqso-eso-cap}
	$\eqso(\eso)$ captures $\spp$ over ordered structures.
\end{proposition}
The previous result connecting fragments from $\# \LL$ with $\shp$ and $\spp$ is the further that we can go by using SO-formulas with free variables. 
So, what happen with other counting complexity classes like $\fp$ and $\fpspace$?
By trying to search for an $\# \LL$ class for capturing $\fp$ one can easily see that this is pointless.
Indeed, by using free SO-variables one will easily capture $\shp$-hard functions (see Section~\ref{sec:syntactic}) and by only using free FO-variables one cannot capture exponential functions definable in $\fp$ (e.g. $2^{n}$).
Therefore, it is impossible to capture classes like $\fp$ or $\fpspace$ if we follow previous approaches like~\cite{SalujaST95}.

Instead, if we move from $\eqso$ we have other options for counting like first- and second-order products. Actually, if we combine $\qfo$ with $\lfp$ this is exactly what we need to capture $\fp$, the class of function that can be computed in~$\ptime$.
\begin{theorem} \label{captfp}
	$\qfo(\lfp)$ captures $\fp$ over ordered structures.
\end{theorem}
To capture $\fp$ one first shows that any formula in $\qfo(\lfp)$ can be evaluated in $\ptime$. 
Indeed, $\lfp$ is a polynomial time logic~\cite{Immerman82,vardi1982complexity} and the sum and product quantifiers can be computed in $\fp$. 
For the other direction, one has to use $\lfp$ to simulate the run of the $\fp$-machine and use the quantitative quantifiers to reconstruct the binary number built in the tape. 
It is important to notice that the alternation between sum and products quantifiers is crucial for capturing $\fp$.
That is, without these operators is difficult to capture $\fp$.

Another natural function complexity class is $\fpspace$~\cite{Ladner89}, the functional version of $\pspace$. 
Of course, for capturing this class one needs a logical core powerful enough, like $\pfp$, for simulating the run of an $\pspace$-machine. 
On the other hand, an $\fpspace$ machine can generate outputs of exponentially many bits. 
For this reason, the second-order sum is not enough for the quantitative layer of a logic for $\fpspace$, that is, one needs to consider second-order products to produce big numbers. 
This leads us to the fragment $\qso(\pfp)$ which naturally captures $\fpspace$. 
\begin{theorem} \label{qso-pfp-cap}
	$\qso(\pfp)$ captures $\fpspace$ over ordered structures.
\end{theorem}
The proof of the previous theorem follows the same lines than the logical characterization of $\fp$: one shows that each function in $\qso(\pfp)$ can be computed in $\fpspace$ and, conversely, the output of an $\fpspace$ can be reconstructed by using $\pfp$ and second-order quantitative quantifiers.

From the proof of the previous theorem a natural question follows: what happen if we use first-order quantitative quantifiers and $\pfp$?
In~\cite{Ladner89}, Ladner also proposes the class $\nfpspace$ of all functions computed by $\pspace$-machines whose output is polynomially bounded. 
Interestingly, if we restrict to FO-quantitative quantifiers we can also capture this class.
\begin{corollary} \label{qfo-pfp-cap}
	$\qfo(\pfp)$ captures $\nfpspace$ over ordered structures.
\end{corollary}

The results in this section validate $\qso$ as the logical framework for extending the theory of Descriptive Complexity to function computational complexity. In the next sections, we study subfragments of $\eqso$ and, moreover, show how to go beyond $\eqso$ to capture other computational classes.

### 2016-12-28 LICS/framework.tex
Primera versión de Sección 4.

We introduce here the logic framework that we use for studying counting complexity classes. 
This logic framework is based on the framework of Weighted Logics~\cite{DrosteG07} (WL) that has been used in the context of weighted automata for studying functions from words (or trees) to semirings. 
We propose here to use the framework of WL over any relational structure and to restrict the semiring to natural numbers. 
The extension to any relational structure will allow us to study general counting complexity classes and the restriction to the natural numbers will simplify the notation in this context (see Section~\ref{sec:previous} below for a more detailed discussion).

@@

Note that the syntax of QSO formulas is divided in two levels. 
The first level is composed by $\so$-formulas over $\R$ (called boolean formulas) and the second level is made by counting operators of addition and multiplication. 
For this reason, the quantifiers in $\so$ (e.g. $\exists x$ or $\exists X$) are called boolean quantifiers and the quantifiers that make use of addition and multiplication (e.g. $\Sigma x$ or $\Pi X$) are called {\em quantitative quantifiers}.
Furthermore, $\Sigma x$ and $\Sigma X$ are called first- and second-order sum, and $\Pi x$ and $\Pi X$ the first- and second-order product, respectively.
This division between boolean and quantitative levels is essential for understanding the difference between the logic and the quantitative part. 
Furthermore, this will allow us later to parametrize both levels of the logic in order to capture different counting complexity classes.

@@

In the next section, we variate $\FF$ and $\LL$ to capture several counting complexity classes. Before this, we devote the next subsection to show the connection of $\qso$ with previous frameworks for defining functions over relational structures.

@@

In this subsection we discuss previous frameworks proposed in the literature and how they differ to our approach.
We start by discussing the connection between $\qso$ and weighted logics (WL)~\cite{DrosteG07}. 
As it was previously discussed, $\qso$ is a fragment of WL.
The real difference is that we restrict the semiring used in WL to natural numbers in order to study counting complexity classes.
Another difference of WL with our approach is that, as far as we know, this is the first paper to study weighted logics over general signature properly to do descriptive complexity of counting classes. 
Previous works on WL usually restrict the signature of the logic to strings, trees, and other specific structures (see \cite{droste2009handbook} for more examples), and did not study the logic over general structures. 
Furthermore, in this paper we propose further extensions for $\qso$ (see Section~\ref{sec:beyond}) which start to differ from previous approaches in WL.

Another approach that resembles $\qso$ are counting logics~\cite{L04,counting-paper-here}. 
Logics with counting operators usually extend $\fo$ with quantifiers that counts how many ways can be satisfied a formula over a second sort (i.e. natural numbers). 
In contrast to our approach, counting operators are usually used for checking boolean properties over structures and not for outputting values (i.e. they do not define a function).
Although there are some papers that uses this operator for counting~\cite{XXX}, they have not used the logic for capturing counting complexity classes.

Finally, the work on~\cite{SalujaST95} and \cite{Gradel91} is of particular interest for our work. 
In~\cite{SalujaST95}, it was proposed to define functions over structures by using open variables on SO-formulas and count how many extensions of the structure satisfy the formula (i.e. by instantiating the open variables). 
Formally, Saluja et. al \cite{SalujaST95} define a family of counting classes $\#\LL$ for each fragment $\LL$ of $\so$. For a formula $\varphi(\bar{x},\bar{X})$ over $\R$, the function $f_{\varphi(x,X)}$ is defined as
$
f_{\varphi}(\A) = \vert \{\langle \bar{a},\bar{A} \rangle\mid \A\models\varphi(\bar{a},\bar{A})\}\vert.
$
for each $\A\in\ostr[\R]$. Then a function $g:\ostr[\R]\to\nat$ is in $\#\LL$ if there exists an formula $\varphi(\bar{x},\bar{X})$ in $\LL$ such that $f = f_{\varphi}$. 
In~\cite{SalujaST95}, it was shown several results for capturing counting complexity classes which are relevant for our work. We discuss and use these results in Sections~\ref{sec:complexity} and~\ref{sec:syntactic}.
Note that for any formula $\varphi(\bar{x},\bar{X})$ it holds that $f_\varphi \equiv \sem{\sa{\bar{X}} \sa{\bar{x}} \varphi(\bar{x},\bar{X})}$, that is, the approach in \cite{SalujaST95} can be seen as a syntactical restriction of $\qso$. 
Thus, the advantage of our approach relies on the flexibility of defining  counting functions by alternating sum with product operators or, moreover, by introducing new quantitative operators (like in Section~\ref{sec:beyond}).
Furthermore, in the next section we show how to capture some interesting classes that could never be captured by following the approach in~\cite{SalujaST95}.

>>>>

We introduce here the logic framework that we use for studying function complexity classes. 
This logic framework is based on the framework of Weighted Logics~\cite{DrosteG07} (WL) that has been used in the context of weighted automata for studying functions from words (or trees) to semirings. 
We propose here to use the framework of WL over any relational structure and to restrict the semiring to natural numbers. 
The extension to any relational structure will allow us to study general function complexity classes and the restriction to the natural numbers will simplify the notation in this context (see Section~\ref{sec:previous} below for a more detailed discussion).

@@

Note that the syntax of QSO formulas is divided in two levels. 
The first level is composed by $\so$-formulas over $\R$ (called boolean formulas) and the second level is made by counting operators of addition and multiplication. 
For this reason, the quantifiers in $\so$ (e.g. $\exists x$ or $\exists X$) are called boolean quantifiers and the quantifiers that make use of addition and multiplication (e.g. $\Sigma x$ or $\Pi X$) are called {\em quantitative quantifiers}.
Furthermore, $\Sigma x$ and $\Sigma X$ are called first- and second-order sum, and $\Pi x$ and $\Pi X$ the first- and second-order product, respectively.
This division between boolean and quantitative levels is essential for understanding the difference between the logic and the quantitative part. 
Furthermore, this will allow us later to parametrize both levels of the logic in order to capture different function complexity classes.

@@

In the next section, we variate $\FF$ and $\LL$ to capture several function complexity classes. Before this, we devote the next subsection to show the connection of $\qso$ with previous frameworks for defining functions over relational structures.

@@

In this subsection we discuss previous frameworks proposed in the literature and how they differ to our approach.
We start by discussing the connection between $\qso$ and weighted logics (WL)~\cite{DrosteG07}. 
As it was previously discussed, $\qso$ is a fragment of WL.
The real difference is that we restrict the semiring used in WL to natural numbers in order to study function complexity classes.
Another difference of WL with our approach is that, as far as we know, this is the first paper to study weighted logics over general signature properly to do descriptive complexity of counting classes. 
Previous works on WL usually restrict the signature of the logic to strings, trees, and other specific structures (see \cite{droste2009handbook} for more examples), and did not study the logic over general structures. 
Furthermore, in this paper we propose further extensions for $\qso$ (see Section~\ref{sec:beyond}) which start to differ from previous approaches in WL.

Another approach that resembles $\qso$ are counting logics~\cite{L04,counting-paper-here}. 
Logics with counting operators usually extend $\fo$ with quantifiers that counts how many ways can be satisfied a formula over a second sort (i.e. natural numbers). 
In contrast to our approach, counting operators are usually used for checking boolean properties over structures and not for outputting values (i.e. they do not define a function).
Although there are some papers that uses this operator for counting~\cite{XXX}, they have not used the logic for capturing function complexity classes.

%\begin{theorem} \label{no-mult}
%	Let $\alpha$ be a $\qso$-formula in some fragment of $\qso$. There is a $\qso$ formula $\alpha'$ in the same fragment such that $\alpha'$ does not have any subformula of the form $(\beta_1 \cdot \beta_2)$ and $\sem{\alpha} = \sem{\alpha'}$.
%\end{theorem}

Finally, the work on~\cite{SalujaST95} and \cite{compton1996logical} is of particular interest for our work. 
In~\cite{SalujaST95}, it was proposed to define functions over structures by using free variables on SO-formulas and count how many extensions of the structure satisfy the formula (i.e. by instantiating the free variables). 
Formally, Saluja et. al \cite{SalujaST95} define a family of counting classes $\#\LL$ for each fragment $\LL$ of $\so$. For a formula $\varphi(\bar{x},\bar{X})$ over $\R$, the function $f_{\varphi(x,X)}$ is defined as
$
f_{\varphi}(\A) = \vert \{\langle \bar{a},\bar{A} \rangle\mid \A\models\varphi(\bar{a},\bar{A})\}\vert.
$
for each $\A\in\ostr[\R]$. Then a function $g:\ostr[\R]\to\nat$ is in $\#\LL$ if there exists an formula $\varphi(\bar{x},\bar{X})$ in $\LL$ such that $f = f_{\varphi}$. 
In~\cite{SalujaST95}, it was shown several results for capturing function complexity classes which are relevant for our work. We discuss and use these results in Sections~\ref{sec:complexity} and~\ref{sec:syntactic}.
Note that for any formula $\varphi(\bar{x},\bar{X})$ it holds that $f_\varphi \equiv \sem{\sa{\bar{X}} \sa{\bar{x}} \varphi(\bar{x},\bar{X})}$, that is, the approach in \cite{SalujaST95} can be seen as a syntactical restriction of $\qso$. 
Thus, the advantage of our approach relies on the flexibility of defining  counting functions by alternating sum with product operators or, moreover, by introducing new quantitative operators (like in Section~\ref{sec:beyond}).
Furthermore, in the next section we show how to capture some interesting classes that could never be captured by following the approach in~\cite{SalujaST95}.
